---
title: 'Part 3: Functional Enrichment'
author: "Stephanie Hickey"
date: "2023-10-25"
output: html_document
---

## Set Up

### Load necessary libraries
```{r, load }
library(tidyverse)
library(DT)
library(patchwork)
library(fgsea)
library(GO.db)
library(RColorBrewer)
conflicted::conflict_prefer(name = "filter", winner = "dplyr")
conflicted::conflict_prefer(name = "select", winner = "dplyr")
conflicted::conflict_prefer(name = "desc", winner = "dplyr")
conflicted::conflict_prefer(name = "rename", winner = "dplyr")

set.seed(1) # to make sure that we get the same results for randomization
```

### Set your project paths in advance 
```{r, set paths}
# tells us which directory we're currently in
# if you opened Rstudio by clicking `RNA-seq_RRM_2023.Rproj`
# you should already be in the `RNA-seq_RRM_2023` directory
getwd() 

# if not use `setwd()` to set the working directory to `RNA-seq_RRM_2023`
# setwd("path/to/RNA-seq_RRM_2023")
project_path <- getwd()

# make a new folder called "part3_results" for the results
# make an object holding folder name
part3_results <- paste0(project_path, "/part3_results") 
# only make the folder if a folder by that name DOES NOT already exist
if(!dir.exists(part3_results)){dir.create(part3_results)} 

```

### Load the data

**Gene association file**

Associates genes with GO terms. There are 17 columns, and you can find the descriptions [here](https://geneontology.org/docs/go-annotation-file-gaf-format-2.1/). We'll only focus on these columns:

  3. `DB Object Symbol`: Gene symbol  
  4. `Qualifier`: Tells how the gene is related to the Gene ontology category   
  5. `GO ID`: Gene ontology category ID   
  
Also interesting:

  7. [`Evidence code`](https://geneontology.org/docs/guide-go-evidence-codes/): Each annotation includes an evidence code to indicate how the annotation to a particular term is supported. Evidence codes fall into several categories:    
      + Inferred from Experiment (EXP)    
      + Inferred from Direct Assay (IDA)    
      + Inferred from Physical Interaction (IPI)    
      + Inferred from Mutant Phenotype (IMP)    
      + Inferred from Genetic Interaction (IGI)   
      + Inferred from Expression Pattern (IEP)    
      + Inferred from High Throughput Experiment (HTP)    
      + Inferred from High Throughput Direct Assay (HDA)    
      + Inferred from High Throughput Mutant Phenotype (HMP)    
      + Inferred from High Throughput Genetic Interaction (HGI)   
      + Inferred from High Throughput Expression Pattern (HEP)    
  9. `Aspect`: The specific ontology the term belongs to. For example, human “cytochrome c” can be described by the *molecular function* oxidoreductase activity, the *biological process* oxidative phosphorylation, and the *cellular component* mitochondrial intermembrane space. If you are only interested in categories belonging to one of these, you can filter for them.   
      + P (biological process)
      + F (molecular function) 
      + C (cellular component)
      
```{r, load gaf}
# gene association file 
gaf <- read_tsv(file = "data/goa_human.gaf", comment = "!", col_names = FALSE)
```

**DESeq2 results**

From Part 2: `part2_results/deseq2_results.txt`

```{r, load deres}
deres <- read_delim(file = "part2_results/deseq2_results.txt", delim = " ") 
```

## Gene set enrichment analysis

Gene Set Enrichment Analysis (GSEA) is a computational method used to determine whether a predefined set of genes, known as a gene set, is significantly overrepresented among genes at the top and bottom of a list of genes ranked by some metric of interest, compared to what would be expected by chance. Check out [Subramanian *et al*. 2005](
https://doi.org/10.1073/pnas.0506580102) for an in depth explanation. 

Here, the gene sets are genes associated with gene ontology terms, and the genes are ranked by the Wald statistic, `stat`, from DESeq2, which is equal to `log2(FoldChange)/standard error`.

### `.gaf` file manipulation 

**What we need:**

A list object where each element is a vector of gene names corresponding to a gene ontology category. 

**How to get there**    

  1. Filter the `Qualifier` column in the gaf file for positive gene/term associations. We don't care about "NOT" associations.   
  2. Select the `Gene` and `GOID` columns.    
  3. Use the `split()` function to split the data frame into a list of data frames by `GOID`    
  4. Use `lapply()` to change each data frame in the list to a vector of genes. Learn more about the `apply()` functions [here](https://ademos.people.uic.edu/Chapter4.html#1_what_are_apply_functions). 
  
```{r, message=FALSE, warning=FALSE}
# Make a list of genes annotated to go terms using the gaf
# filter out "NOT" relationships
gaf_list <- 
  gaf %>%
  filter(!grepl("NOT", X4)) %>%  # filter out "NOT" relationships
  dplyr::select(X5, X3) %>%      # another package has a function called `select()`
  rename(GOID = "X5",            # rename() changes the column names
         Gene = "X3") %>%
  distinct() %>%
  dplyr::select(GOID, Gene) %>%
  split(as.factor(.$GOID))       # split the df into a list by GOID
                                 # because `split()` is not a tidyverse function
                                 # we need to used .$ to refer to the GOID column

gaf_list[[1]]                    # access the first element of the list
gaf_list[[2]]                    # access the second element of the list 
length(gaf_list)                 # how many list elements?
head(names(gaf_list))            # names of the list elements

# make a list of vectors containing the genes instead of a list of data frames
class(gaf_list[[1]])

go_list <- lapply(gaf_list, function(x) {y = x$Gene; y})

class(go_list[[1]])

# look at the first few list elements
head(go_list)

# names carry over from gaf_list
head(names(go_list)) # names carry over from gaf_list

```

### GSEA with `fgsea`

We need to make a vector of all of the genes from the DESeq2 results ranked by the Wald statistic, `stat`. 
```{r}
# make a vector of genes ranked by stat
DEranks <- 
  deres %>%
  filter(!is.na(padj)) %>%       # remove genes with no calculated padj
  select(Gene, stat) %>%         # select Gene and stat cols
  arrange(stat) %>%              # ranks the genes by stat
  deframe()                      # turns the data frame into a named vector

head(DEranks)
tail(DEranks)

# perform GSEA analysis
fgseaRes <- fgsea(pathways = go_list, 
                  stats    = DEranks,
                  minSize  = 10,      # smallest goterm to consider
                  maxSize  = 500)     # largest goterm to consider

head(fgseaRes)
```

**Column descriptions**

  * `pathway` – name of the pathway as in ‘names(pathway)‘; here it's the GOID    
  * `pval` – an enrichment p-value   
  * `padj` – a BH-adjusted p-value   
  * `log2err` – the expected error for the standard deviation of the P-value logarithm    
  * `ES` – enrichment score, reflects the degree to which the genes in a gene set are overrepresented at the top or bottom of the entire ranked list of genes,
      + $ES>0$ means the genes are enriched among up-regulated genes   
      + $ES<0$ means the genes are enriched among down-regulated genes   
  * `NES` – enrichment score normalized to mean enrichment of random samples of the same size   
  * `size` – size of the pathway after removing genes not present in ‘names(DEranks)‘   
  * `leadingEdge` – leading edge genes that drive the enrichment   

### Reformat the results table 

Use the GO.db package to get the descriptions that matching the GOIDs. Later, we'll add this to the results table using `left_join()`   
```{r}
# extract all go term descriptions from the GO.db package
goterms <- as.data.frame(Term(GOTERM))

# change column name
colnames(goterms) <- "GOTerm"

# make the rownames into a column called `GOID`
goterms$GOID <- rownames(goterms)

head(goterms)
```

Change the format of the `leadingEdge` from a list of character vectors to a vector of strings with `sapply()` and `toString()` 
```{r}
# leading edge is a list of character vectors, where each row is a list element
class(fgseaRes$leadingEdge)

fgseaRes$leadingEdge[[1]]
# we can only print cells in our data frame with length = 1
length(fgseaRes$leadingEdge[[1]])

# to print it as a tab delimited file, we need to change the column into one character vector
# so we'll put all of the genes into one string
# using toString() and sapply()

toString(fgseaRes$leadingEdge[[1]])
length(toString(fgseaRes$leadingEdge[[1]]))

# another example
fgseaRes$leadingEdge[[2]]
length(fgseaRes$leadingEdge[[2]])

toString(fgseaRes$leadingEdge[[2]])
length(toString(fgseaRes$leadingEdge[[2]]))

# format the results table
fgseaReform <- 
  fgseaRes %>%
  mutate(leadingEdge = sapply(leadingEdge, toString),      # change leadingEdge from list to character vector
         leadingEdgeSize = str_count(leadingEdge, "\\S+"), # count the number of leadingEdge genes
         Regulation = case_when(                           # add a `Regulation` column
           NES < 0 ~ "down-regulated",
           NES > 0 ~ "up-regulated")) %>%
         rename(GOID = "pathway") %>%             
         left_join(goterms) %>%                            # add a column with the go term descriptions using left_join()
         relocate(GOTerm, .after = GOID)  %>%              # move GOTerm to after GOID
         arrange(padj)

head(fgseaReform)
         
write_delim(fgseaReform, file = paste(part3_results,  "fgsea_results.csv", sep = "/"))

# make interactive table
fgseaTab <- 
  fgseaReform %>%
  filter(padj < .05) %>%
  dplyr::select(-pval, 
                -log2err,
                -ES,
                -NES,
                -leadingEdgeSize,
                -size) %>%
  mutate(Regulation = as.factor(Regulation),
         padj = format(padj, digits=3)) 
  
datatable(fgseaTab, rownames = FALSE, filter = "top")
```

## Plot results

  * Check how many GO terms are significantly enriched for up- or down-regulated genes    
  * Make separate bubble plots for terms enriched for up- or down-regulated genes with:
      + The top 10 terms by padj on the y-axis   
      + The `NES` on the x-axis   
      + A "bubble" for each term colored by `-log10(padj)` and sized by `leadingEdgeSize`   

```{r, fig.width=13, fig.height=5}
# How many significant terms are there?
fgseaReform %>%        
  filter(padj < .05) %>%      # filter for significantly enriched terms
  group_by(Regulation) %>%    # group the rows by `Regulation`
  tally()                     # count the rows in each group

# make a color palette 
sequential_pal <- colorRampPalette(brewer.pal(n = 9, name ="YlOrRd"))(20)
# remove the lightest colors because they are hard to see
sequential_pal_dark <- sequential_pal[5:20]

# find the terms of interest
down_oi <- 
  fgseaReform %>%
    filter(Regulation == "down-regulated") %>%                
    slice_min(order_by=padj, n=10) %>%
    pull(GOTerm)

  up_oi <- 
  fgseaReform %>%
    filter(Regulation == "up-regulated") %>%                
    slice_min(order_by=padj, n=10) %>%
    pull(GOTerm)

# find the smallest and largest leadingEdgeSize among the terms of interest
small_edge =
  fgseaReform %>%
  filter(GOTerm %in% c(down_oi, up_oi)) %>%
  pull(leadingEdgeSize) %>%
  min()

small_edge

large_edge =
  fgseaReform %>%
  filter(GOTerm %in% c(down_oi, up_oi)) %>%
  pull(leadingEdgeSize) %>%
  max()

large_edge

# find the highest and lowest -log10(padj) among the terms of interest
high_adj =
  fgseaReform %>%
  filter(GOTerm %in% c(down_oi, up_oi)) %>%
  mutate(log_padj = -log10(padj)) %>%
  pull(log_padj) %>%
  max() %>%
  ceiling() # rounds up to closest whole number

high_adj

low_adj =
  fgseaReform %>%
  filter(GOTerm %in% c(down_oi, up_oi)) %>%
  mutate(log_padj = -log10(padj)) %>%
  pull(log_padj) %>%
  min() %>%
  floor() # rounds down to closest whole number

low_adj

# plot
plot_down_terms <- 
  fgseaReform %>%
    filter(Regulation == "down-regulated") %>%            # filter for terms enriched with down-regulated genes            
    slice_min(order_by=padj, n=10) %>%                    # select the terms with the 10 lowest padj
    mutate(GOTerm = fct_reorder(GOTerm, desc(NES))) %>%   # order the terms by NES for the plot
    ggplot(aes(x=NES,                 
               y=GOTerm, 
               size=leadingEdgeSize,                      # size of the bubble = leadingEdgeSize 
               color = -log10(padj))) +                   # color the bubble by -log10(padj)
      geom_point() +                                      # make a dot (bubble) plot
      xlim(0,-4) +
      lims(size = c(small_edge,large_edge)) +              # force the size limits to between small_edge and large_edge 
      scale_color_gradientn(colours = sequential_pal_dark, # use the sequential_pal_dark to color the bubbles
                            limits= c(low_adj,high_adj)) + # force the color key to go from low_adj to high_adj 
      labs(title="Gene Ontology Enrichment",                           
           subtitle = "down-regulated genes") +
      xlab("Normalized Enrichment Score") +
      theme_bw()

plot_up_terms <- 
  fgseaReform %>%
    filter(Regulation == "up-regulated") %>%
    slice_min(order_by=padj, n=10) %>%
    mutate(GOTerm = fct_reorder(GOTerm, NES)) %>%
    ggplot(aes(x=NES, y=GOTerm, 
               size=leadingEdgeSize, 
               color = -log10(padj))) +
      geom_point() + 
      xlim(0,4) +
      lims(size = c(small_edge,large_edge)) +  
      scale_color_gradientn(colours = sequential_pal_dark,
                            limits= c(low_adj,high_adj)) +
      labs(title="Gene Ontology Enrichment",                           
           subtitle = "up-regulated genes") +
      xlab("Normalized Enrichment Score") +
      theme_bw()

# patchwork magic to combine the two plots
both = (plot_down_terms | plot_up_terms) + plot_layout(guides = 'collect') 
both

ggsave(both, file = paste(part3_results, "goterm_bubble.png", sep = "/"), width=13, height=5)
```


